Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex
Progress and operational semantics.

\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{calc}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{color}
\usepackage{xspace}
\usepackage{array}
\usepackage{multirow}
\usepackage{varwidth}
\usepackage{tabularx}  % for 'tabularx' environment and 'X' column type
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{xcolor,colortbl}
\usepackage{graphicx}

@@

\newcommand{\restore}{\texttt{restore}}

@@

\ADComment{Think about whether ``restore'' is best name.}

@@

We use two worked examples---work stealing and breadth-first graph
search---to provide background on the OpenCL programming model and
irregular data-parallel algorithms.

@@

\paragraph{OpenCL}
An OpenCL program is divided into \emph{host} and \emph{device}
components.  A host application runs on the CPU and launches one or
more \emph{kernels} that run on accelerator devices---GPUs in the
context of this paper.  A kernel is writen in OpenCL C, based on C99.
All threads executing a kernel execute the same entry function with
identical arguments, but a thread may query its unique thread id in
order to access different data or follow different control flow paths
to other threads.

@@

We describe our cooperative kernels programming model as an extension
to the industry-standard OpenCL programming model for heterogeneous
many-core architectures.  However, the cooperative kernels concept is
more general, and could be applied to extend other GPU programming
models, such as CUDA and HSA.

We describe the semantics of cooperative kernels
(Section~\ref{sec:semantics}), then summarise the important
differences between programming with regular vs.\ cooperative kernels
(Section~\ref{sec:programmingguidelines}) and the non-functional
properties that a developer of a cooperative kernel, and the
implementer of a scheduler for cooperative kernels, should strive for
(Section~\ref{sec:nonfunctional}).  The semantics of cooperative
kernels has been guided by the applications that we have studied
(described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
have taken particular semantic decisions where other reasonable
alternatives were possible (Section~\ref{sec:semanticalternatives}).
We conclude the section by discussing the backwards-compatible nature
of cooperative kernels with respect to blocking algorithms designed to
run on current GPU platforms
(Section~\ref{sec:backwardscompatibility}).

@@

\ADComment{A natural exit does not decrement $M$.}

As with a regular OpenCL kernel, a cooperative kernel is launched by
the host application, with a specified number of workgroups, each
consisting of a specified number of threads.  The kernel's behaviour
is described by a kernel function, written in OpenCL C but using the
additional language extensions that we present below, and a number of
parameters to the kernel function are passed by the host application.

The parameters to a cooperative kernel are \emph{immutable}, as if
they were qualified with the \texttt{const} keyword (though pointer
parameters can refer to mutable data).

\ADComment{Perhaps remove this because a driver-based approach could handle this.}
A regular OpenCL kernel can be executed by a multi-dimensional grid of
workgroups, called an ND-range.  In our current proposal, the ND-range
for a cooperative kernel must be one-dimensional.  \ADComment{Say why?
  And mention that none of the applications that we deem would benefit
  from cooperative kernels need a multi-dimensional ND-range.}

As in regular OpenCL, each workgroup can be multi-dimensional.

\paragraph{Restored variables}

A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new \restore qualifier.  By
annotating a variable with \restore, the programmer indicates
that when new workgroups are spawned via $\offerfork$, the threads in
these workgroups should obtain a well-defined initial value for the
variable, obtained from the workgroup that executed $\offerfork$.  We
detail the semantics for this when we describe $\offerfork$ below.

\ADComment{Remove?}
Only private memory variables can be annotated with \restore,
which may include pointers to local or global memory so long as the
pointer variables themselves are stored in private memory.  In
particular, variables and arrays in local memory cannot be annotated
with \restore.  \ADComment{The preceding is what we have
  implemented, but actually it is pretty clear how we could handle
  local memory restoration if we wanted to; it's just that none of our
  examples needed it.}

@@

\item for each variable $v$ annotated with \restore, every new
  thread's copy of $v$ is initialised to the value that thread 0 in
  workgroup $m$ held for $v$ at the point of the $\offerfork$ call.

@@

@@

\ADComment{Make the point here that if an irregular algorithm is
  written as a cooperative kernel then it will execute just as safely
  as it does today on current hardware if: (1) the \restore
  annotation and $\offerkill$/$\offerfork$ calls are defined away to
  nothing, (2) $\resizingglobalbarrier$ is redefined to
  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
  that assumes the occupancy-bound execution model is provided.  If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}

@@

@@

>>>>

\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{calc}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{color}
\usepackage{xspace}
\usepackage{array}
\usepackage{multirow}
\usepackage{varwidth}
\usepackage{tabularx}  % for 'tabularx' environment and 'X' column type
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{xcolor,colortbl}
\usepackage{graphicx}
\usepackage{mathpartir}

@@

\newcommand{\transmit}{\mathsf{transmit}}

@@

@@

We use two examples of irregular data-parallel algorithms---a work stealing queue and breadth-first graph
search---to introduce the industry-standard OpenCL programming model on which we base cooperative kernels.

@@

\paragraph{OpenCL kernels}

An OpenCL program is divided into \emph{host} and \emph{device}
components.  A host application runs on the CPU and launches one or
more \emph{kernels} that run on accelerator devices---GPUs in the
context of this paper.  A kernel is writen in OpenCL C, based on C99.
All threads executing a kernel execute the same entry function with
identical arguments, but a thread may query its unique thread id in
order to access different data or follow different control flow paths
to other threads.

@@

We describe our cooperative kernels programming model as an extension
to OpenCL.  However, the cooperative kernels concept is more general,
and could be applied to extend other GPU programming models, such as
CUDA and HSA.

We describe the semantics of cooperative kernels
(Section~\ref{sec:semantics}), then summarise the important
differences between programming with regular vs.\ cooperative kernels
(Section~\ref{sec:programmingguidelines}) and the non-functional
properties that a developer of a cooperative kernel, and the
implementer of a scheduler for cooperative kernels, should strive for
(Section~\ref{sec:nonfunctional}).  The semantics of cooperative
kernels has been guided by the applications that we have studied
(described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
might have taken different, equally reasonable, semantic decisions (Section~\ref{sec:semanticalternatives}).
We conclude the section by discussing the backwards-compatible nature
of cooperative kernels with respect to blocking algorithms designed to
run on current GPU platforms
(Section~\ref{sec:backwardscompatibility}).

@@

As with a regular OpenCL kernel (see Section~\ref{sec:background}), a
cooperative kernel is launched by the host application.  The host
application passes parameters to the kernel and specifies a desired
number of workgroups, each consisting of a specified number of
threads.  The parameters to a cooperative kernel are \emph{immutable}, as if
they were qualified with the \texttt{const} keyword (though pointer
parameters can refer to mutable data).

A cooperative kernel is written in OpenCL C plus the following
language extensions: $\transmit$, a
new qualifier on the local variables of a thread; $\offerkill$ and
$\offerfork$, the key functions that enable cooperative scheduling;
$\globalbarrier$ and $\resizingglobalbarrier$, two primitives to allow
inter-workgroup synchronisation.

We now explain the semantics of these language extensions, after which
we explain how the work stealing and graph traveral algorithms of
Section~\ref{sec:background} can be turned into cooperative kernels
that use the extensions.

\paragraph{Transmitted variables}

A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new $\transmit$ qualifier.  By
annotating a variable with $\transmit$, the programmer indicates that
when a workgroup spawns new workgroups by calling $\offerfork$, the
workgroup should transmit the current value of this variable to serve
as an initial value for the new workgroups.  We detail the semantics
for this when we describe $\offerfork$ below.

\ADComment{Remove?}
Only private memory variables can be annotated with $\transmit$,
which may include pointers to local or global memory so long as the
pointer variables themselves are stored in private memory.  In
particular, variables and arrays in local memory cannot be annotated
with $\transmit$.  \ADComment{The preceding is what we have
  implemented, but actually it is pretty clear how we could handle
  local memory restoration if we wanted to; it's just that none of our
  examples needed it.}

@@

\item for each variable $v$ annotated with $\transmit$, every new
  thread's copy of $v$ is initialised to the value that thread 0 in
  workgroup $m$ held for $v$ at the point of the $\offerfork$ call.

@@

\paragraph{Examples}

\ADComment{TODO}

@@

\ADComment{Make the point here that if an irregular algorithm is
  written as a cooperative kernel then it will execute just as safely
  as it does today on current hardware if: (1) the $\transmit$
  annotation and $\offerkill$/$\offerfork$ calls are defined away to
  nothing, (2) $\resizingglobalbarrier$ is redefined to
  $\globalbarrier$, (3) a standard implementation of $\globalbarrier$
  that assumes the occupancy-bound execution model is provided.  If the kernel uses blocking features---the global barrier, mutexes, etc., then it may not execute safely due to the possibility of unfair scheduling, but it will execute no less safely than it would if the cooperative kernels style were not used.}

@@

\ADComment{A natural exit does not decrement $M$.  We haven't made a point of this, and I don't think we should right now, but we should remember to think this through carefully.}

@@

\section{Semantics}

Thread state: $(l, \mathit{ss})$, where $l \in L$ is the thread's
local state, and $\mathit{ss}$ is a semi-colon-separated sequence of
program statements that the thread remains to execute.

Workgroup state: $((l_1, \mathit{ss}_1), \dots, (l_d,
\mathit{ss}_d))$, where each $(l_i, \mathit{ss}_i)$ is a thread state.

Kernel state: an $N$-tuple $(\sigma, (w_1, w_2, \dots, w_M, \bot,
\dots, \bot))$, where each $w_i$ is a workgroup state, which are
followed by $N-M$ occurrences of $\bot$ to indicate absent workgroups.

We treat thread-level semantics abstractly by assuming a transition
relation $(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l',
\mathit{ss}'))$ that is defined unless $\mathit{ss} =
\mathsf{special}(); \mathit{tt}$, where $\mathit{special}$ is one of
$\offerkill$, $\offerfork$, $\globalbarrier$ or
$\resizingglobalbarrier$.

To abstractly account for the interaction between global barrier
synchronization and the GPU memory model, we assume a $\mathsf{sync}$
that can be applied to the shared state $\sigma$, or to the shared
state and a thead local state.  In the first case,
$\mathsf{sync}(\sigma)$ returns an updated shared state that reflects
all in-flight memory stores having completed.  In the second case,
$\mathsf{sync}(\sigma, l)$ returns a thread local state reflecting all
in-flight memory loads issued by the thread having completed.

\begin{figure*}
\begin{center}

\[
\inferrule{
w_i(j) = (l, \mathit{ss})
\\
(\sigma, (l, \mathit{ss})) \rightarrow_{\tau} (\sigma', (l', \mathit{ss}'))
\\
w_i' = w_i[j \mapsto (l', \mathit{ss}')]
}
{
(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma', (\dots, w_i', \dots))
}
(\textsc{Thread-Step})
\]

\medskip

\[
\inferrule{
\forall j \;.\; w_i(j) = (l_j, \offerkill();\mathit{ss})
\\
w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
}
{
(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
}
(\textsc{Kill-No-Op})
\]

\medskip

\[
\inferrule{
\forall j \;.\; w_M(j) = (l_j, \offerkill();\mathit{ss})
}
{
(\sigma, (\dots, w_{M-1}, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_{M-1}, \bot, \bot, \dots, \bot))
}
(\textsc{Kill})
\]

\medskip

\[
\inferrule{
\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
\\
w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
}
{
(\sigma, (\dots, w_i, \dots)) \rightarrow (\sigma, (\dots, w_i', \dots))
}
(\textsc{Fork-No-Op})
\]

\medskip

\[
\inferrule{
\forall j \;.\; w_i(j) = (l_j, \offerfork();\mathit{ss})
\\
w_i' = ((l_1, \mathit{ss}), \dots, (l_d, \mathit{ss}))
\\
k \in [0, N - M]
\\
\forall a \in [1, k] \;.\; w_{M+a} = ((l_1, \mathit{ss}), \dots, (l_1, \mathit{ss}))
}
{
(\sigma, (\dots, w_i, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (\dots, w_i', \dots, w_M, w_{M+1}, \dots, w_{M+k}, \bot, \dots, \bot))
}
(\textsc{Fork})
\]

\medskip

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \globalbarrier();\mathit{ss})
\\
\forall i \;.\;\forall j\;.\; l_{i,j}' = \mathsf{sync}(\sigma, l_{i,j})
\\
\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}', \mathit{ss})
\\
\sigma' = \mathsf{sync}(\sigma)
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma', (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Barrier})
\]

\medskip

\[
\inferrule{
\forall i \;.\;\forall j\;.\;w_i(j) = (l_{i,j}, \resizingglobalbarrier();\mathit{ss})
\\
\forall i \;.\;\forall j\;.\;w_i'(j) = (l_{i,j}, \offerkill(); \globalbarrier(); \offerfork();\mathit{ss})
}
{
(\sigma, (w_1, \dots, w_M, \bot, \dots, \bot)) \rightarrow (\sigma, (w_1', \dots, w_M', \bot, \dots, \bot))
}
(\textsc{Resizing-Barrier})
\]

\end{center}

\caption{Abstract operational semantics for our cooperative kernels language extensions}

\end{figure*}
